---
title: "Getting Started with mfvar2"
author: "Sam Bartlett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mfvar2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE # Set to TRUE when data is available
)
```

## Introduction

The `mfvar2` package implements mixed frequency Bayesian vector autoregression (MF-BVAR) for macroeconomic forecasting, with particular focus on Swiss National Bank (SNB) data. The methodology combines:

- **Mariano & Murasawa (2003)** for mixed frequency modeling
- **Schorfheide & Song (2013)** for MF-VAR forecasting
- **Durbin & Koopman (2001)** for state-space methods
- **Giannone et al. (2008)** for empirical Bayes hyperparameter selection
- **Litterman/Doan** for Minnesota prior

## Basic Workflow

### Step 1: Load and Prepare Data

```{r data_prep}
library(mfvar2)

# Prepare SNB data
data_prepared <- prepare_data_snb(
  data_sources = c("path/to/cpi.csv", "path/to/gdp.csv"),
  quarterly_vars = c("gdp", "investment"),
  flow_vars = c("gdp"), # GDP is flow-type
  start_date = "2000-01-01",
  end_date = "2024-12-31",
  log_transform = c("cpi", "gdp", "investment"),
  difference = TRUE, # Apply differencing based on unit root tests
  verbose = TRUE
)

# Inspect prepared data
print(data_prepared)
summary(data_prepared$metadata)
```

### Step 2: Tune Hyperparameters

```{r hyperparameter_tuning}
# Tune Minnesota prior tightness λ via marginal data density
hyper_optimal <- tune_minnesota_hyper(
  data_prepared = data_prepared,
  p = 6, # 6 lags for monthly data
  lambda_grid = c(0.05, 0.1, 0.2, 0.5, 1.0, 2.0, 3.0, 5.0),
  refine_continuous = TRUE,
  verbose = TRUE,
  seed = 123
)

# View results
print(hyper_optimal)

# Plot marginal data density over grid
plot_mdd_grid(hyper_optimal)
```

### Step 3: Estimate MF-BVAR

```{r estimation}
# Estimate Bayesian VAR with Gibbs sampling
posterior <- estimate_mf_bvar(
  data_prepared = data_prepared,
  p = 6,
  hyperparameters = hyper_optimal,
  n_draws = 5000,
  burnin = 1000,
  thinning = 2,
  verbose = TRUE,
  seed = 456
)

# Inspect posterior
print(posterior)

# Posterior mean coefficients
print(posterior$A_mean)

# Posterior mean residual covariance
print(posterior$Sigma_mean)
```

### Step 4: Generate Forecasts

```{r forecasting}
# Generate predictive distributions
forecasts <- forecast_mf_bvar(
  posterior = posterior,
  horizon_months = 12,
  n_sim = 1000,
  seed = 789
)

# Inspect forecasts
print(forecasts)

# Monthly forecasts for CPI
print(forecasts$forecasts_monthly["cpi", , ])

# Quarterly forecasts for GDP
print(forecasts$forecasts_quarterly[["gdp"]])
```

### Step 5: Visualize with Fan Charts

```{r fan_charts, fig.width=7, fig.height=5}
# Fan chart for CPI
plot_fan_chart(
  forecast = forecasts,
  variable = "cpi",
  intervals = c(0.5, 0.7, 0.9)
)

# Fan chart for GDP
plot_fan_chart(
  forecast = forecasts,
  variable = "gdp",
  intervals = c(0.5, 0.7, 0.9)
)
```

### Step 6: Rolling Window Evaluation

```{r rolling_window}
# Out-of-sample forecast evaluation
eval_results <- rolling_forecasts_mf_bvar(
  data_prepared = data_prepared,
  p = 6,
  window_type = "expanding",
  min_window = 120, # 10 years minimum
  horizons = c(1, 3, 6, 12),
  n_draws = 2000,
  burnin = 500,
  reuse_hyperparams = TRUE,
  include_ar2 = TRUE,
  verbose = TRUE,
  seed = 999
)
```

### Step 7: Compare with AR(2) Benchmarks

```{r evaluation}
# Evaluate forecast accuracy
accuracy <- evaluate_forecast_accuracy(
  mfvar_forecasts = eval_results$mfvar_forecasts,
  ar2_forecasts = eval_results$ar2_forecasts,
  actuals = eval_results$actuals
)

# View results
print(accuracy)

# RMSE ratios (values < 1 mean MF-BVAR is better)
print(accuracy$RMSE_ratio)
```

## Advanced Usage

### Custom Hyperparameters

```{r custom_hyper}
# Manually specify hyperparameters
custom_hyper <- list(
  lambda = 0.5,
  theta = 1.0,
  kappa_cross = 0.5
)

posterior_custom <- estimate_mf_bvar(
  data_prepared = data_prepared,
  p = 6,
  hyperparameters = custom_hyper,
  n_draws = 5000,
  seed = 123
)
```

### Rolling vs Expanding Windows

```{r rolling_types}
# Expanding window (default)
eval_expanding <- rolling_forecasts_mf_bvar(
  data_prepared = data_prepared,
  p = 6,
  window_type = "expanding",
  min_window = 120
)

# Rolling window with fixed width
eval_rolling <- rolling_forecasts_mf_bvar(
  data_prepared = data_prepared,
  p = 6,
  window_type = "rolling",
  rolling_width = 120,
  min_window = 120
)
```

## Methodological Notes

### Mixed Frequency Handling

- **Monthly variables**: Directly observed each month
- **Stock quarterly variables**: Observed at quarter-end (Mar, Jun, Sep, Dec)
- **Flow quarterly variables**: Sum of three monthly values (Mariano-Murasawa aggregation)

### Minnesota Prior

- **Prior mean**: Own first lag = 1 (random walk), others = 0
- **Prior variance**: $\text{Var}(A_{ij}^{(l)}) = \frac{\lambda^2 \kappa_{ij}^2}{l^\theta} \frac{\sigma_i^2}{\sigma_j^2}$
- **Hyperparameters**:
  - $\lambda$: overall tightness (optimized via MDD)
  - $\theta$: lag decay (default: 1)
  - $\kappa$: cross-equation shrinkage (default: 0.5)

### Gibbs Sampler

Three-step iteration:

1. **Draw states**: Carter-Kohn simulation smoother
2. **Draw coefficients**: Equation-by-equation Normal posterior
3. **Draw variances**: Inverse-Gamma posterior

## References

1. Mariano, R. S., & Murasawa, Y. (2003). A new coincident index of business cycles based on monthly and quarterly series. *Journal of Applied Econometrics*, 18(4), 427-443.

2. Schorfheide, F., & Song, D. (2013). Real-time forecasting with a mixed-frequency VAR. *NBER Working Paper*.

3. Durbin, J., & Koopman, S. J. (2001). *Time Series Analysis by State Space Methods*. Oxford University Press.

4. Giannone, D., Lenza, M., & Primiceri, G. E. (2008). Priors for the long run. *Manuscript*, Universite Libre de Bruxelles.

5. Litterman, R. B. (1986). Forecasting with Bayesian vector autoregressions—five years of experience. *Journal of Business & Economic Statistics*, 4(1), 25-38.
