---
title: "MIDAS Regressions using Swiss Data"
author: "Minna"
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r, show=FALSE, output=FALSE}
rm(list = ls())
library(midasr)
library(ggplot2)
library(kofdata)
library(lubridate)
library(dplyr)
library(forecast)
```

## Import Data

```{r}
# to render
data_raw <- read.csv("../../data/data_quarterly.csv")
metadata <- read.csv("../../data/metadata_quarterly.csv")
# to run
data_raw <- read.csv("submission/data/data_quarterly.csv")
metadata <- read.csv("submission/data/metadata_quarterly.csv")

metadata <- metadata[, c(1,8)]

# transform to ts object
data <-  data_raw |>
   mutate(
    # TODO: alternatively, with inflation?
    # take real valued gdp
    gdp_growth = 100 * (log(rvgdp) - dplyr::lag(log(rvgdp)))
  ) |>
  filter(!is.na(gdp_growth)) |>
  select(date, gdp_growth)
# all y data

y_ext <- ts(data$gdp_growth, start = c(1992, 1), end =  c(2027, 4), frequency=4) 
# y - without forecasts
# because we diff later, go one step forward to make up for loss of 1 val for x
y <- window(y_ext, start = c(1992, 1), end =  c(2025, 3))  # quarterly
# monthly time series -> kof (swiss) economic barometer
baro <- get_time_series("ch.kof.barometer")$ch.kof.barometer
```

#### account for differences in publication

for more information, see `publication_times.md`

```{r}
# helper to align monthly indicator with quarterly GDP release lag
align_monthly_to_quarter <- function(y_ts, x_ts, shift_months, m = 3) {
  if (shift_months < 0) stop("shift_months must be >= 0")
  total_months <- length(x_ts)
  total_quarters <- length(y_ts)
  if (shift_months >= total_months) stop("shift_months too large for available data")
  
  # lead monthly series by the number of months that are already observed when GDP is released
  x_shifted <- stats::lag(x_ts, -shift_months)
  
  available_quarters <- floor((total_months - shift_months) / m)
  complete_months <- available_quarters * m
  if (available_quarters < 1) stop("Not enough data left after applying the shift.")
  
  # keep only months/quarters that remain observable after the shift
  x_shifted <- stats::window(
    x_shifted,
    end = stats::time(x_shifted)[complete_months]
  )
  y_shifted <- stats::window(
    y_ts,
    end = stats::time(y_ts)[available_quarters]
  )
  
  list(y = y_shifted, x = x_shifted, quarters = available_quarters)
}

# try these different publication gaps (0, 1, or 2 months)
month_shift <- c(0, 1, 2)
selected_shift <- 0  # average GDP release lag from publication_times.md
m <- 3  # number of high-frequency observations per quarter
```


#### diff and align monthly series
```{r}
# align monthly series so that after differencing it starts at the first month
# of the first quarter in y
y_start <- start(y)                 # c(year, quarter)
y_start_year <- y_start[1]
y_start_quarter <- y_start[2]
first_month_of_quarter <- (y_start_quarter - 1) * 3 + 1
prev_month <- first_month_of_quarter - 1
if (prev_month == 0) {
  prev_month <- 12
  prev_year <- y_start_year - 1
} else {
  prev_year <- y_start_year
}

# fetch one extra month before the first quarter, then difference and realign
xx0 <- window(baro, start = c(prev_year, prev_month))
x <- diff(xx0)
# after diff, both have the same start -> suitable length for midas regression
x <- window(x, start = c(y_start_year, first_month_of_quarter))

# re-apply the publication lag alignment using the differenced monthly series
# TODO: try again with 1, 0, compare...
aligned_series <- align_monthly_to_quarter(
  y_ts = y,
  x_ts = x,
  shift_months = selected_shift,
  m = m
)
y <- aligned_series$y
x <- aligned_series$x
```


#### check for stationarity
```{r}
# TODO: check for seasonality as well
# TODO: why acf not whole numbers
acf(x)
pacf(x)
```

significant at lag 1.

```{r}
acf(diff(y))
pacf(diff(y))
```

If y is decomposed, it's white noise

## Train/Test Split

This is just preliminary until general CV is ready.

```{r}
n <- length(y)
n_val <- ceiling(0.1*n)

# Split y data
y_train <- window(y, end = time(y)[n - n_val])
y_val <- window(y, start = time(y)[n - n_val + 1])

# Get the end quarter of y_train
y_train_end_quarter <- time(y_train)[length(y_train)]
y_train_end_year <- floor(y_train_end_quarter)
y_train_end_q <- round((y_train_end_quarter - y_train_end_year) * 4) + 1

# Calculate when that quarter's GDP was published (2 months after quarter end)
# Q1 ends in March -> published in May (month 5)
# Q2 ends in June -> published in August (month 8)  
# Q3 ends in September -> published in November (month 11)
# Q4 ends in December -> published in February (month 2 of next year)
publication_months <- c(5, 8, 11, 2)  # months when GDP is published (2 months after quarter end)
publication_month <- publication_months[y_train_end_q]

# If Q4, publication is in February of next year
if (y_train_end_q == 4) {
  x_train_end_year <- y_train_end_year + 1
  x_train_end_month <- publication_month
} else {
  x_train_end_year <- y_train_end_year
  x_train_end_month <- publication_month
}

# x_train ends at the publication month of the last y_train quarter
# This ensures we only use monthly data that was available when that quarter's GDP was published
x_train <- window(x, end = c(x_train_end_year, x_train_end_month))

# Verify x_train length matches y_train (accounting for m months per quarter)
# After alignment, x_train should have length(y_train) * m months
if (length(x_train) != length(y_train) * m) {
  # Adjust to match exactly
  required_x_train_length <- length(y_train) * m
  if (length(x_train) > required_x_train_length) {
    # Trim excess months
    x_train <- window(x_train, end = time(x_train)[required_x_train_length])
  } else {
    # This shouldn't happen if alignment worked correctly, but handle it
    warning("x_train shorter than expected. Check alignment.")
  }
}

# Get the start quarter of y_val to determine when it was published
y_val_start_quarter <- time(y_val)[1]
y_val_start_year <- floor(y_val_start_quarter)
y_val_start_q <- round((y_val_start_quarter - y_val_start_year) * 4) + 1

# Calculate when that quarter's GDP was published (2 months after quarter end)
# Q1 ends in March -> published in May (month 5)
# Q2 ends in June -> published in August (month 8)  
# Q3 ends in September -> published in November (month 11)
# Q4 ends in December -> published in February (month 2 of next year)
y_val_publication_month <- publication_months[y_val_start_q]

# If Q4, publication is in February of next year
if (y_val_start_q == 4) {
  x_val_start_year <- y_val_start_year + 1
  x_val_start_month <- y_val_publication_month
} else {
  x_val_start_year <- y_val_start_year
  x_val_start_month <- y_val_publication_month
}

# x_val should have exactly length(y_val) * m months
required_x_val_length <- length(y_val) * m
x_val_end_month <- x_val_start_month + required_x_val_length - 1
x_val_end_year <- x_val_start_year
while (x_val_end_month > 12) {
  x_val_end_month <- x_val_end_month - 12
  x_val_end_year <- x_val_end_year + 1
}

x_val <- window(x, start = c(x_val_start_year, x_val_start_month),
                end = c(x_val_end_year, x_val_end_month))

# Verify alignment
cat("y_train: ", start(y_train), "to", end(y_train), "(", length(y_train), "quarters)\n")
cat("x_train: ", start(x_train), "to", end(x_train), "(", length(x_train), "months, should be", length(y_train) * m, ")\n")
cat("y_val: ", start(y_val), "to", end(y_val), "(", length(y_val), "quarters)\n")
cat("x_val: ", start(x_val), "to", end(x_val), "(", length(x_val), "months, should be", length(y_val) * m, ")\n")

trend_train <- 1:length(y_train)
trend_val <- (length(y_train)+1) : (length(y_train)+length(y_val))
```

#### check data

```{r}
tail(y_train)
tail(x_train)
```

## Fit Models

#### beispiel, show fmls()

```{r}
# print(fmls(x=x_train, k=2, m=3))
head(y_train)
# mls(y_train, k=2, m=1)
# mls(y_train, k=1, m=1) 
# mls(y_train, k=1:2, m=1) 
# TODO: mls(y_train, k=1:2, m=1)
# mls(y_train, k=2, m=1)

```

#### benchmark models

```{r}
# AR(2) model
ar1 <- stats::arima(y_train, order = c(2, 0, 0))
summary(ar1)

rw <- forecast::rwf(y_train, h = length(y_val), drift = TRUE)
summary(rw)
```

<!-- TODO: use forecast package cross validation -->


#### Midas w/ AR(1) and trend 


##### to determine proper lag of y data
```{r}
acf(y)
pacf(y)
```

We can see, lag 2 significant 

```{r midas}
# check length of fmls matrix 

m <- 3 # needs to be length(x)/m=length(y) so, m = 3

f <- fmls(x_train, k=2, m=3, nealmon)
length(f)

# with trend 
m_ar1_t <- midas_r(
  y_train ~ 
  trend_train + # incl. trend 
  mls(y_train, k=1:2, m=1) +
  # mls(y_train, k=2, m=1) +  # only lag 2
  fmls(x_train, k=2, m=3), # max lag 1 year # add nealmon sometime
  start = list(x_train = rep(0, 3))
  )

summary(m_ar1_t)
```

#### Midas w/ AR(1) and no trend 

```{r}
# without trend
m_ar1 <- midas_r(
  y_train ~ 
  mls(y_train, k=1:2, m=1) +
  # mls(y_train, k=2, m=1) + # only lag 2
  fmls(x=x_train, k=2, m=3), # max lag 1 year # add nealmon sometime
  start = list(x_train = rep(0, 3))
  )

summary(m_ar1)
```


### Unrestricted Midas w/ trend  and AR(1)

```{r}
# try out midas_nlpr 
# mu_it <- midas_nlpr(y_train ~ trend_train + mls(y_train,1,1) + fmls(x_train,2,3), start=list(x_train = rep(0, 3)))
# forecast::summary(mu_it)
```


## Forecast 

#### Midas w/ AR(1) and trend
```{r}
h <- 4 # one year point forecasts

# here need to constrict x_val as well so that length(x_val)/m = length(y_val)
# w/ trend
m_ar1_t_forecast <- midasr::forecast(
  m_ar1_t, 
  # need to name them the same as above, if not, doesn't work
  newdata=list(x_train=x_val, trend_train=trend_val), 
  h=4, 
  method='dynamic' # dynamic, aka uses y_hat t for forecasting y hat t+1 -> because only 1 lag used for y (see model spec above)
  )

summary(m_ar1_t_forecast)
# from now on, only actual values is forecast
m_ar1_t_forecast <- m_ar1_t_forecast$mean
```

#### Midas w/ AR(1) and no trend

```{r}
# h <- 4 # one year point forecasts

# here need to constrict x_val as well so that length(x_val)/m = length(y_val)
# w/ trend
m_ar1_forecast <- midasr::forecast(
  m_ar1, 
  # need to name them the same as above, if not, doesn't work
  newdata=list(x_train=x_val, trend_train=trend_val), 
  h=4, 
  method='dynamic' # dynamic, aka uses y_hat t for forecasting y hat t+1 -> because only 1 lag used for y (see model spec above)
  )

summary(m_ar1_forecast)
# from now on, only actual values is forecast
m_ar1_forecast <- m_ar1_forecast$mean
length(m_ar1_forecast)
```

### unrestricted midas forecast

```{r}
# mu_it_forecast <- midasr::forecast(
#   mu_it, 
#   # need to name them the same as above, if not, doesn't work
#   newdata=list(x_train=x_val, trend_train=trend_val), 
#   h=4, 
#   method='dynamic' # dynamic, aka uses y_hat t for forecasting y hat t+1 -> because only 1 lag used for y (see model spec above)
#   )

# summary(mu_it_forecast)
# mu_it_forecast <- mu_it_forecast$mean
```

### benchmark forecasts
```{r benchmark forecasts}
# define horizon based on other models
h <- length(m_ar1_t_forecast)

ar_forecast <- forecast(ar1, h=h)$mean
rw_forecast <- rw$mean[seq_len(h)]
print(length(rw_forecast))
print(length(ar_forecast))
# TODO: length 14 but rest is length 13
```

### create Plot DF

```{r df_plot, warnings=FALSE}
h <- length(m_ar1_t_forecast)
df_plot <- data.frame(
  # TODO: change time to actual date
  time = (T - 19):(T + h),
  actual = c(tail(y_train, 20), rep(NA, h)),
  actual_post = c(rep(NA, 20), tail(y_val, h)),
  m_ar1_t_forecast = c(rep(NA, 20), as.numeric(m_ar1_t_forecast)),
  m_ar1_forecast = c(rep(NA, 20), as.numeric(m_ar1_forecast)),
  ar_forecast = c(rep(NA, 20), as.numeric(ar_forecast)),
  rw_forecast =c(rep(NA, 20), as.numeric(rw_forecast))
  # ,
  # mu_it_forecast =c(rep(NA, 20), as.numeric(mu_it_forecast))
)

df_plot |> select(-c(actual)) |> slice_tail(n=10)
```

### Plot

```{r plot}
ggplot(df_plot, aes(x = time)) +
  geom_line(aes(y = actual), color = "black", size = 1, linetype = "solid") +
  geom_line(aes(y = actual_post), color = "black", size = 1, linetype = "dashed") +
  geom_line(aes(y = m_ar1_forecast, color = "Simple MIDAS"), size = 1) +
  geom_line(aes(y = m_ar1_t_forecast, color = "Trend MIDAS"), size = 1) +
  geom_line(aes(y = ar_forecast, color = "AR(1)"), size = 1) +
  geom_line(aes(y = rw_forecast, color = "Random Walk"), size = 1) +
  # geom_line(aes(y = mu_it_forecast, color = "MU trend"), size = 1) +
  labs(
    title = "MIDAS Forecast Comparison",
    y = "Y",
    color = "Model"
  ) +
  theme_minimal(base_size = 14) +
  geom_vline(xintercept = T, linetype = "dashed")

ggsave(filename = "../../output/midas_forecast_plot_shifted.png", width = 10, height = 6, dpi = 300)

```

### Evaluation

```{r eval}
# Get forecast period data only
eval_df <- tail(df_plot, h) |> 
  select(-c('actual')) |>
  mutate(
    # Calculate squared errors for each model
    se_trend = (actual_post - m_ar1_t_forecast)^2,
    se_simple = (actual_post - m_ar1_forecast)^2,
    se_ar = (actual_post - ar_forecast)^2,
    se_rw = (actual_post - rw_forecast)^2
    # ,
    # se_mu_it =  (actual_post - mu_it_forecast)^2
  )

# Calculate MSE for each model
mse_results <- data.frame(
  Model = c("MIDAS with trend", "MIDAS without trend", "AR(1)", "Random Walk"
  # , "MU with trend"
  ),
  MSE = c(
    mean(eval_df$se_trend, na.rm = TRUE),
    mean(eval_df$se_simple, na.rm = TRUE),
    mean(eval_df$se_ar, na.rm = TRUE),
    mean(eval_df$se_rw, na.rm = TRUE)
    #,
    # mean(eval_df$se_mu_it, na.rm = TRUE)
  ),
  RMSE = c(
    sqrt(mean(eval_df$se_trend, na.rm = TRUE)),
    sqrt(mean(eval_df$se_simple, na.rm = TRUE)),
    sqrt(mean(eval_df$se_ar, na.rm = TRUE)),
    sqrt(mean(eval_df$se_rw, na.rm = TRUE))
    #,
    # sqrt(mean(eval_df$se_mu_it, na.rm = TRUE))
  )
)

# Show both the detailed errors and summary statistics
print("Detailed errors by time period:")
# print(eval_df |> select(-c('time')))
print("\nSummary MSE/RMSE by model:")
print(mse_results)
write.csv(mse_results, file = "../../output/mse_results_shifted.csv")
```

```{r}
# Random walk w/ drift -> auto.arima generates if no AR or MA component needed 
# rw_drift <- forecast::auto.arima(y_train) # ARIMA (0,1,0)
# summary(rw_drift)
```

So for now, the `auto.arima` fitted model, which is a random walk, see output  with trend performs better than the model without trend.